---
title: "Practical Machine Learning - Week 2"
author: "Saul Lugo"
date: "January 11, 2016"
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, chache=TRUE)
```

#Splitting Data, Plotting Predictors and Training Models
The following are examples of how to split the data set in training and testing sets, how to train the model and how to plot the predictors to analyze the relationship between the predictors and the outcome.

#Loading the Data
Is this example, the ISLR packages is used. This package has a dataset of Wages in the US.

```{r loading_data}
require(ISLR); require(ggplot2); require(caret);
data(Wage)
head(Wage)
summary(Wage)
```

#Splitting the Data into Training and Test set

```{r splitting_data}
inTrain <- createDataPartition(y = Wage$wage, p = 0.7, list = FALSE)
training <- Wage[inTrain,]
testing <- Wage[-inTrain,]
dim(training); dim(testing)
```

#Plotting Predictors vs Outcome

```{r plotting}
#Plotting several predictors vs the outcome
featurePlot(x = training[,c("age", "education", "jobclass")], y = training$wage, plot="pairs")

#Plotting one variable vs outcome and adding a second variable in the colour
qplot(age, wage, colour = jobclass,data=training)

#Add regression smoothers
qq <- qplot(age, wage, colour=education, data=training)
qq + geom_smooth(method="lm", formula = y ~ x)

#cut2, making factors (Hmisc package)
require(Hmisc)
#Splitting the wage variable into groups of quantiles
cutWage <- cut2(training$wage, g=3)
table(cutWage)
#Making a boxplot to see the three different wage groups we created before
p1 <- qplot(cutWage, age, data=training, fill=cutWage, geom = c("boxplot"))
p1

#Boxplots with points overlayed
#If the jitter plot shows a lot of the points inside the boxplots that mean that the boxplots are
#actually representative of the data, so any trend one might observes might be true.
#On the contrary if only a few points are shown inside the boxplots, the trend might not be that representative
p2 <- qplot(cutWage, age, data = training, fill = cutWage, geom = c("boxplot","jitter"))
#grid.arrange(p1, p2, ncol=2)
p2

#One can make also tables
t1 <- table(cutWage,training$jobclass)
t2 <- table(cutWage,training$race)
t3 <- table(cutWage,training$education)
t1; t2; t3

#One can also use prop.table to get the proportion on each group
prop.table(t2,1)

#Also, one can do Density Plots
qplot(wage, colour=education, data=training, geom="density")
```

#Preprocessing Predictor Values

```{r preprocessing}
library(caret)
library(kernlab)
data(spam)
inTrain <- createDataPartition(y=spam$type, p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
hist(training$capitalAve,main="Capital in a Row in the emails of the dataset",xlab="ave. capital run length")
mean(training$capitalAve)
sd(training$capitalAve)
```
It can be observed that this variable is highly skewed. So it can be improved by preprocessing.

##Preprocessing by Normalization (Standarization)
To standarize a variable one must substract the mean and divide the result by the SD of the variable:

```{r standarize}
trainCapAve <- training$capitalAve
trainCapAveS <- (trainCapAve - mean(trainCapAve))/sd(trainCapAve)
round(mean(trainCapAveS),4)
round(sd(trainCapAveS),4)
```

Also, the function **preProcess** can be used for standarization:

```{r standarize_function}
preObj <- preProcess(training[,-58],method=c("center","scale"))
trainCapAveS <- predict(preObj,training[,-58])$capitalAve
mean(trainCapAveS)
sd(trainCapAveS)
```

If the standarization is done in the test set, the mean and the SD of the training set must be use still. However, after standarized the test set variable the mean of the standarized variable will not be exactly zero neither the SD will be exactly one:

```{r standarization_over_testing_set}
testCapAveS <- predict(preObj,testing[,-58])$capitalAve
mean(testCapAveS)
sd(testCapAveS)
```

The **preProcess** function can be passed directly to the **train** function:

```{r training_model}
set.seed(32343)
modelFit <- train(type ~ ., data=training, preProcess=c("center","scale"),method="glm")
modelFit
```

Other transformation available is the **BoxCox** transformation:

```{r box_cox_transformation}
preObj <- preProcess(training[,-58],method=c("BoxCox"))
trainCapAveS <- predict(preObj, training[,-58])$capitalAve
par(mfrow=c(1,2)); hist(trainCapAveS); qqnorm(trainCapAveS)
```

##Preprocessing Imputing Missing Values

If the dataset has missing values, those can be imputing using **K-nearest neighbor's imputation** algorithm:

```{r knearest}
#Make some values NA
training$capAve <- training$capitalAve
selectNA <- rbinom(dim(training)[1], size=1, prob=0.05)==1
training$capAve[selectNA] <- NA

#Impute and Standarize
preObj <- preProcess(training[,-58],method="knnImpute")
capAve <- predict(preObj, training[,-58])$capAve

#Standarize true values
capAveTruth <- training$capitalAve
capAveTruth <- (capAveTruth-mean(capAveTruth))/sd(capAveTruth)
```

#Creating Covariates (or Features)

In case that one of the predictors is a factor variable, it is better to transform that variable into dummy variables. Prediction algoritms work better with dummy variables than with factor variables:

```{r dummy_variables}
library(ISLR); library(caret); data(Wage);
inTRain <- createDataPartition(y=Wage$wage, p=0.7, list=FALSE)
training <- Wage[inTrain,]; testing <- Wage[-inTrain,];

#converting the jobclass variable from a qualitative variable to a quantivative variable
#using dummyVars function from the caret package
table(training$jobclass)
dummies <- dummyVars(wage ~ jobclass, data=training)
head(predict(dummies, newdata=training))
```

##Removing zero covariates

In order to detect those variables that has close to none variability, and therefore are not useful for prediction, one can use the **nearZerVar** function:

```{r nzv_function}
nsv <- nearZeroVar(training, saveMetrics=TRUE)
nsv
```

##Principal Componen Analysis (PCA)

```{r pca}
library(caret)
library(kernlab)
data(spam)
inTrain <- createDataPartition(y=spam$type, p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]

#PCA using the basic pacakge funtion prcomp
typeColor <- ((spam$type=="spam")*1 + 1) #clasifies each point for coloring as spam or ham
prComp <- prcomp(log10(spam[,-58]+1)) #the log10 is to make the variables look more "normal"
plot(prComp$x[,1],prComp$x[,2],col=typeColor,xlab="PC1",ylab="PC2")

#PCA using the caret package
preProc <- preProcess(log10(spam[,-58]+1),method="pca",pcaComp=2)
spamPC <- predict(preProc,log10(spam[,-58]+1))
plot(spamPC[,1],spamPC[,2],col=typeColor)

#One can fit a model with the Training set and the principal componets
preProc <- preProcess(log10(training[,-58]+1),method="pca",pcaComp=2)
trainPC <- predict(preProc,log10(training[,-58]+1))
modelFit <- train(training$type ~ .,method="glm",data=trainPC)
modelFit
confusionMatrix(training$type,predict(modelFit,trainPC))

#Now in the test dataset
testPC <- predict(preProc,log10(testing[,-58]+1)) #one must use the same preProc obj calculated for the training
confusionMatrix(testing$type,predict(modelFit,testPC)) #one must also use the same model fitted for the trining

#Another way of train the model and use the PCA in the preprocessing at the same time:
modelFit <- train(training$type ~ .,method="glm",preProcess="pca",data=training)
confusionMatrix(training$type,predict(modelFit,training))
confusionMatrix(testing$type,predict(modelFit,testing))
```

#Predicting with Regression Models
Linear Models using the caret package:

```{r lm_caret}
library(MASS)
data(faithful); set.seed(333)
inTrain <- createDataPartition(y=faithful$eruptions, p=0.5, list=FALSE)
trainFaith <- faithful[inTrain,]; testFaith <- faithful[-inTrain,]
head(trainFaith)

modFit <- train(eruptions ~ waiting,data=trainFaith,method="lm")
summary(modFit$finalModel)
plot(trainFaith$waiting,trainFaith$eruptions)
lines(trainFaith$waiting,modFit$finalModel$fitted,col="blue")
```

#Predicting with Regression Models with Multiple variable

```{r mv_lm}
library(ISLR); library(ggplot2); library(caret)
data(Wage); Wage <- subset(Wage,select=-c(logwage))
summary(Wage)

#Splitting the data
inTrain <- createDataPartition(y=Wage$wage,p=0.7,list=FALSE)
training <- Wage[inTrain,]; testing <- Wage[-inTrain,]
dim(training); dim(testing)

#fitting a linear model
modFit <- train(wage ~ age + jobclass + education, method="lm", data=training) #we fit the model on the training data

fitMod <- modFit$finalModel
print(modFit)

#Plotting the fitted values vs the residuals
plot(fitMod,1,pch=19,cex=0.5,col="#00000010")
#You can see there are some outliers in the residuals
#Let't plot fitted vs residuals and add the race variable and see if that variable explains the outliers in the residuals
qplot(fitMod$fitted, fitMod$residuals, colour=race, data=training)
#You see that the outliers might be explained by the race variable
#also plot the residuals accross the dataset and see if there is a patron. The residual should be random
plot(fitMod$residuals,pch=19)

#Predicted versus true values in the test set
pred <- predict(modFit, testing)
qplot(wage, pred, colour=year,data=testing)