---
title: "Practical Machine Learning - Week3"
author: "Saul Lugo"
date: "January 22, 2016"
output: pdf_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, chache=TRUE)
```
#Predicting with Trees
The basic algorithm for predicting with trees is the following:

1) Start with all the variables in one group
2) Find the variable/split that best separates the outcomes
3) Divide the data into two groups ("leaves") on that split
4) Within each split, find the variable that best separates the outcome
5) Continue until the groups are too small or sufficiently pure

##Example with the Iris dataset

```{r iris_prediction}
data(iris)
library(ggplot2)
library(caret)
names(iris)

table(iris$Species)

#Create the data partition
inTrain <- createDataPartition(y=iris$Species,p=0.7,list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
dim(training); dim(testing)

#Exploring the data
qplot(Petal.Width,Sepal.Width,colour=Species,data=training)

#Training the model (a tree model)
modFit <- train(Species ~ ., method="rpart", data = training)
modFit_PredTree <- modFit
print(modFit$finalModel)

#Plotting the model
plot(modFit$finalModel, uniform=TRUE,main="Classification Tree")
text(modFit$finalModel, use.n=TRUE, all=TRUE, cex=0.8)

##A better plot
library(rattle)
fancyRpartPlot(modFit$finalModel)

#Now predicting the outcome in the testing set
prediction <- predict(modFit,newdata=testing)
prediction
CM_PredTree <- confusionMatrix(testing$Species,prediction)
CM_PredTree
```

##Notes on predicting with trees

1) Classification trees are non-linear models
        - They use interactions between variables
        - Data transformation might be less important
        - Trees can also be use for regression problems (continious outcome)
2) There are several options for building trees in R:
        - party, rpart in the caret package
        - tree, out of the carect package

#Bagging with the Caret Package

Bagging stands for **Bootstrap Aggregating**.

The following is an example of Bagging using the caret package:

```{r bagging}
library(ElemStatLearn); data(ozone,package="ElemStatLearn")
library(caret)
head(ozone)
predictors = data.frame(ozone=ozone$ozone)
temperature = ozone$temperature
treebag <- bag(predictors,temperature,B=10,bagControl = bagControl(fit = ctreeBag$fit, predict = ctreeBag$pred, aggregate = ctreeBag$aggregate))
treebag$fit
plot(ozone$ozone, temperature, col="lightgrey", pch=19)
#Plotting the prediction 1 of the treebag
points(ozone$ozone, predict(treebag$fits[[1]]$fit,predictors),pch=19,col="red")
#Plotting the prediction 9 of the treebag
points(ozone$ozone, predict(treebag$fits[[9]]$fit,predictors),pch=19,col="green")
#Plotting the prediction 5 of the treebag
points(ozone$ozone, predict(treebag$fits[[5]]$fit,predictors),pch=19,col="black")
#Plotting the aggregated predictions of the treebag
points(ozone$ozone, predict(treebag,predictors),pch=19, col="blue")
```

We can see that the aggregated prediction (the blue dots in the plot) is the closest to the real values of temperature.

Also we can verify that the aggregated prediction has the lowest **RSME** (Root Mean Squared Error):

```{r calculating_rsme}
#The following is the RSME of the aggregated prediction:
sqrt(sum((ozone$temperature-predict(treebag,predictors))^2)/111)
#The next 10 are the RSME of each of the fitted predictions. One can see that the lowest one is the RSME of the aggregated prediction.
sqrt(sum((ozone$temperature-predict(treebag$fits[[1]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[2]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[3]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[4]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[5]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[6]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[7]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[8]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[9]]$fit,predictors))^2)/111)
sqrt(sum((ozone$temperature-predict(treebag$fits[[10]]$fit,predictors))^2)/111)
```

#Random Forrest

The **Random Forrest** algorithm is similar to bagging. It bootstraps samples of the data and built a tree. At each split of the tree it bootstrap a set of the varibles (of the predictors). Then, the algorithm also creates many trees following the same logic.

For the prediction, each set of predictors is passed through each tree and then the final answer is the average of all the answers from each tree.

RF is a hightly accurate algorithm, however it can lead to overfitting. Therefore, cross-validation must be used in order to detect the overfitting.

The following is an example of RF algorithm in R:

```{r rf_example}
data(iris); library(ggplot2)
library(caret)
inTrain <- createDataPartition(y=iris$Species,p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
modFit_RF <- train(Species ~ .,data=training,method="rf",prox=TRUE)
modFit_RF

#Comparing the RF model with the Prediction Tree Model
modFit_PredTree

#I can check one of the specific trees:
getTree(modFit_RF$finalModel,k=2)

#Predicting over the testing set:
pred <- predict(modFit_RF,testing)
confusionMatrix(testing$Species,pred)

#Comparing the confusion Matrices of RF model and the pred tree model:

CM_PredTree
```

Comparing the confusion matrix of the random forest algorithm over the testing set with that of the prediction tree, we can see that the RF is more accurate.

#Boosting

Boosting algorithm takes several classifiers weigths and averages them in order to obtain a better one. It can use trees, glms, RF trees, etc. In R, in the **caret** packages, there are several options for boosting:

- **gbm** - boosting with trees
- **mboost** - model based boosting
- **ada** - statistical boosting based on additive logistic regression
- **gamBoost** - for boosting generalized additive models

In the next example we will use boosting prediction in the same problem of predicting the flower Species:

```{r boosting_prediction}
data(iris); library(ggplot2)
library(caret)
inTrain <- createDataPartition(y=iris$Species,p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
MF_Boosting <- train(Species ~ .,method="gbm",data=training,verbose=FALSE)
pred <- predict(MF_Boosting,testing)
confusionMatrix(testing$Species,pred)
```

#Model Based Prediction
The basic ideas of model based prediction are:

- Assume that data follow a probabilistic model
- Use Bayes's theorem to identify optimal classifiers

There are several algorithms for Model Based Prediction:

- **Linear Discriminant Analysis - lda:** the discrimination function is a multivariate Gaussian with the same covariances
- **Quadratic Discriminant Analysis - qda:** the discrimination function is multivariate Gaussian with different covariances
- **Model Based Prediction**: assumes more complicated versions of the covariance matrix
- **Naive Bayes - nb:** assumes independence between features model building

In the following example we compare the Linear Discriminant Analysis algorithm with the Naive Bayes:

```{r lda_nb}
modlda <- train(Species ~ .,data=training,method="lda")
modnb <- train(Species ~ .,data=training,method="nb")
plda <- predict(modlda,testing)
pnb <- predict(modnb,testing)
confusionMatrix(testing$Species,plda)
confusionMatrix(testing$Species,pnb)
table(plda,pnb)
equalPred <- plda==pnb
qplot(Petal.Width,Sepal.Width,colour=equalPred,data=testing)
```

In the plot we can see that the values in which the predictions desagree are in the boundary between two classes.

#Quiz-3
#Question-1:
Load the cell segmentation data from the AppliedPredictiveModeling package using the commands:

library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret)
1. Subset the data to a training set and testing set based on the Case variable in the data set.

2. Set the seed to 125 and fit a CART model with the rpart method using all predictor variables and default caret settings.

3. In the final model what would be the final model prediction for cases with the following variable values:

a. TotalIntench2 = 23,000; FiberWidthCh1 = 10; PerimStatusCh1=2

b. TotalIntench2 = 50,000; FiberWidthCh1 = 10;VarIntenCh4 = 100

c. TotalIntench2 = 57,000; FiberWidthCh1 = 8;VarIntenCh4 = 100

d. FiberWidthCh1 = 8;VarIntenCh4 = 100; PerimStatusCh1=2

```{r question1}
library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret)
training <- subset(segmentationOriginal,Case=="Train")
testing <- subset(segmentationOriginal,Case=="Test")
set.seed(125)
model_CART <- train(Class ~.,data=training,method="rpart")
library(rattle)
fancyRpartPlot(model_CART$finalModel)
```

Analyzing the Plot of the tree growth by the CART model we can see that the answer must be:

a.- PS
b.- WS
c.- PS
d.- Not possible to predict


#Question-2
If K is small in a K-fold cross validation is the bias in the estimate of out-of-sample (test set) accuracy smaller or bigger? If K is small is the variance in the estimate of out-of-sample (test set) accuracy smaller or bigger. Is K large or small in leave one out cross validation?

The bias is smaller and the variance is smaller. Under leave one out cross validation K is equal to one.

The bias is larger and the variance is smaller. Under leave one out cross validation K is equal to the sample size.

The bias is larger and the variance is smaller. Under leave one out cross validation K is equal to two.

The bias is smaller and the variance is bigger. Under leave one out cross validation K is equal to one.

As a reference for this question I read the following link
[Cross-validation in Wikipedia] https://en.wikipedia.org/wiki/Cross-validation_(statistics)

When k=n (the sample size) k-fold cross-validation is the same as leave-one-out cross-validation.

Also, the k is small you get more bias and less variance, so the answer is:

**The bias is larger and the variance is smaller. Under leave one out cross validation K is equal to the sample size.**

#Question-3

Load the olive oil data using the commands:

library(pgmm)
data(olive)
olive = olive[,-1]
(NOTE: If you have trouble installing the pgmm package, you can download the -code-olive-/code- dataset here: olive_data.zip. After unzipping the archive, you can load the file using the -code-load()-/code- function in R.)

These data contain information on 572 different Italian olive oils from multiple regions in Italy. Fit a classification tree where Area is the outcome variable. Then predict the value of area for the following data frame using the tree command with all defaults

newdata = as.data.frame(t(colMeans(olive)))
What is the resulting prediction? Is the resulting prediction strange? Why or why not?

2.783. There is no reason why this result is strange.

4.59965. There is no reason why the result is strange.

0.005291005 0 0.994709 0 0 0 0 0 0. The result is strange because Area is a numeric variable and we should get the average within each leaf.

2.783. It is strange because Area should be a qualitative variable - but tree is reporting the average value of Area as a numeric variable in the leaf predicted for newdata

```{r question3}
library(pgmm)
data(olive)
olive = olive[,-1]
head(olive)
system.time(modFit <- train(Area ~., data=olive, method="rpart"))
newdata = as.data.frame(t(colMeans(olive)))
newdata
predict(modFit,newdata)
fancyRpartPlot(modFit$finalModel)
summary(olive$Area)
```

Examining the tree model plot we can see that the prediction is correct (2.783). In my opinion the result is not extrange because in the dataset "Area" is not a qualitative variable but a quantitative one.

#Question-4

Load the South Africa Heart Disease Data and create training and test sets with the following code:

library(ElemStatLearn)
data(SAheart)
set.seed(8484)
train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
trainSA = SAheart[train,]
testSA = SAheart[-train,]
Then set the seed to 13234 and fit a logistic regression model (method="glm", be sure to specify family="binomial") with Coronary Heart Disease (chd) as the outcome and age at onset, current alcohol consumption, obesity levels, cumulative tabacco, type-A behavior, and low density lipoprotein cholesterol as predictors. Calculate the misclassification rate for your model using this function and a prediction on the "response" scale:

missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}
What is the misclassification rate on the training set? What is the misclassification rate on the test set?

```{r question4}
library(ElemStatLearn)
data(SAheart)
set.seed(8484)
train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
trainSA = SAheart[train,]
testSA = SAheart[-train,]
set.seed(13234)
head(trainSA)
modFit <- train(chd ~ age + alcohol + obesity + tobacco + typea + ldl,data=trainSA,method="glm",family="binomial")

missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}

missClass(trainSA$chd,predict(modFit,trainSA))
missClass(testSA$chd,predict(modFit,testSA))
```

Therefore, the misclassification error on the training set is 0.27 and on the test set is 0.31

#Question-5

Load the vowel.train and vowel.test data sets:

library(ElemStatLearn)
data(vowel.train)
data(vowel.test)
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit a random forest predictor relating the factor variable y to the remaining variables. Read about variable importance in random forests here: http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr The caret package uses by default the Gini importance.

Calculate the variable importance using the varImp function in the caret package. What is the order of variable importance?

The order of the variables is:

x.10, x.7, x.9, x.5, x.8, x.4, x.6, x.3, x.1,x.2

The order of the variables is:

x.1, x.2, x.3, x.8, x.6, x.4, x.5, x.9, x.7,x.10

The order of the variables is:

x.10, x.7, x.5, x.6, x.8, x.4, x.9, x.3, x.1,x.2

The order of the variables is:

x.2, x.1, x.5, x.6, x.8, x.4, x.9, x.3, x.7,x.10

```{r question5}
library(ElemStatLearn)
data(vowel.train)
data(vowel.test)
summary(vowel.train)
vowel.train$y <- as.factor(vowel.train$y)
vowel.test$y <- as.factor(vowel.test$y)
set.seed(33833)
library(caret)
modFit <- train(y ~., data=vowel.train,method="rf")
modFit
varImp(modFit)
```

